Incorporating periodic variability in hidden Markov models for animal movement

### 2016-03-02 paper3.Rnw
working on paper

Clustering time-series data into discrete groups can improve
prediction as well as providing insight into the nature of underlying,
unobservable states of the system. However, temporal variation in the
probability that individuals occupy or move between groups can obscure
such signals. We use finite mixture and hidden Markov models, two
standard clustering techniques, to model high-resolution hourly
movement data from Florida panthers. Allowing for temporal
heterogeneity in transition probabilities, a straightforward but
rarely explored model extension, resolves some shortcomings of current
modeling frameworks and clarifies the behavioural patterns of
panthers. More generally, incorporating previously neglected
structure in statistical models can allow researchers to more
accurately identify models of appropriate complexity.

@@

\cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as 6 for individuals with a large amount
of available data); for practical reasons, they restricted their
detailed analysis to models with only 3 underlying states.  In
contrast, most studies using HMM have chosen the
number of underlying states \emph{a priori}, typically using either two
\citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,fryxell_multiple_2008}, or three states
\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,kerk2015hidden}. As \cite{kerk2015hidden} comment, behavioural
repertoires with more than three distinct states are difficult to
interpret, one possible reason that other authors have not adopted
\citeauthor{kerk2015hidden}'s model-based approach to
identifying the number of latent states.

Our second goal, therefore, is to explore whether
\citeauthor{kerk2015hidden}'s results might be driven at
least in part by structural problems with the HMM, i.e. the
assumption of temporally homogeneous behaviour.  When large data sets
are available, information-theoretic model selection methods will
typically choose complex, highly parameterized models; when there is
only one way in which models can become more complex (e.g. by
increasing the number of latent states), complexity that is
present in the data but not accounted for in the model (e.g. spatial
or temporal heterogeneity) can be misidentified as other forms of
complexity.  We predict that increasing volumes of data will
increasingly lead researchers who are accustomed to fitting 
small models to sparse data into such traps.  We examine whether
allowing for diurnal variation in the Florida panther data leads to
selection of models with smaller numbers of latent states; we also
fit models to simulated data with varying numbers of latent states
and degrees of temporal heterogeneity to test our conjecture that
heterogeneity can be misidentified as behavioural complexity.

@@

\cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
an extension of HMM that permits explicit modelling of dwell times 
\citep{langrock_flexible_2012}, considering both Poisson and
negative binomial distributions for dwell times.  As shown 
by \cite{kerk2015hidden} (Figure S3b, top row, middle panel),
the estimated shape parameter of the negative binomial dwell time
distribution was typically close to 1 ($\approx 0.4-1.6$; confidence
intervals were not calculated), implying that a geometric distribution
(i.e., negative binomial with shape=1) might be adequate. In turn,
this suggests that we might not lose much accuracy by reverting to 
a simpler HMM framework, which makes precisely this assumption.

@@

\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent states.
In general, BIC values decreased as the number of states
increased from 
3 to 6 states,
suggesting that the 6-state model was
favoured statistically; however,
the authors used 3-state models in most
of their analyses for ease of biological interpretation.
We follow \cite{kerk2015hidden} in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the ``true'' number 
of underlying movement states.  Using BIC also simplifies 
evaluation of model selection procedures; 
it is easier to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\citep{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \citep{BurnhamAnderson1998};
we suspect that the qualitative conclusions stated here
for BIC-optimality will carry over to analyses using AIC instead.

@@

In a HMM, the joint likelihood of \emph{emissions} (i.e., direct observations) $\mathbf Y = \obs_{1},..., \obs_{T}$ and a hidden state sequence $\mathbf{Z}, \state_{t} \in \{1,...,n\}, t=1,...,T$, given model parameters {$\boldsymbol \theta$} and covariates $\mathbf{X}_{1:T} =\cov_{1},..., \cov_{T}$, can be written as:
\begin{equation}
\begin{split}
P(\mathbf{Y}_{1:T},\mathbf{Z}_{1:T}|\boldsymbol \theta,\mathbf{X}_{1:T}) & =
P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \times \\
&  \quad \prod\limits_{k=2}^{T}P(\state_{k} | \state_{k-1}, \cov_{k})P(\obs_{k} | \state_{k},\cov_{k})
\end{split}
\label{eq:HMMlik}
\end{equation}

The model contains three distinct components:

\begin{description}
\item[Initial probability] $P(\state_{1} = i | \cov_{1}) P(\obs_{1} | \state_{1}, \cov_{1} )$: the probability of state $i$ at time $t=1$ where the covariate is $\cov_{1}$; the vector of observation densities $\obs_{1}$ conditions on covariates $\cov_{1}$ and state $\state_{1}$.

@@

\item[Emission probability] $P(\obs_{k} | \state_{k},\cov_{k})$: a vector of observation density $\obs_{k}$ condition on covariates $\cov_{k}$ at state $\state_{k}$ at time $t=k$.  
\end{description}

@@

We considered three different transition models for diurnal variation in behaviour, incorporating hour-of-day as a covariate following the general approach of \cite{morales_extracting_2004} of incorporating covariate dependence in the transition matrix.

\begin{description}
\item[Multiple block transition] Here we assume piecewise-constant transition probabilities. The transition probability $\pi_{ij}$ is a function of time (hour of day), where it is assigned to one of $M$ different time blocks:
$$
\eta_{ij}(t) = \sum_{m=1}^{M}a_{ijm} \delta_{m=t}
$$
where $a_{ijm}$ are parameters, and $\delta_{m=t}$ is a Kronecker delta ($\delta_{m=t}=1$ for the time block at the corresponding time $t$, and 0 otherwise).  
\item[Quadratic transition model] We assume the elements of the linear predictor
are quadratic functions of hour. The quadratic model is not diurnally continuous, i.e. there is no constraint that forces $\eta_{ij}(0)=\eta_{ij}(24)$; imposing a diurnal continuity constraint would collapse the model to a constant.
$$
\eta_{ij}(t) = b_{ij1}+b_{ij2}\left(\frac{t}{24}\right)+b_{ij3}\left(\frac{t}{24}\right)^{2}
$$
\item[Sinusoidal transition model] A sinusoidal model with a period of 24 hours is identical in complexity to the quadratic model, but automatically satisfies the diurnal continuity constraint.
$$
\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos\left(\frac{2\pi t}{24}\right)+b_{ij3}\sin\left(\frac{2\pi t}{24}\right)
$$
\end{description}

Model complexity and number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity. See supplementary material Table 1.

@@

We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers with the most data ($>  $ observations). (1) Comparing BICs to the optimal-BIC model within each type of transition complexity ($\dbic$ = BIC - min(BIC)). 
(2) Compare average step-length by hour of day for the observed data and for data simulated from the models.
(3) Compare temporal autocorrelations for the observed data and for data simulated from the models. For the second and third metrics (hourly average and ACF), we simulated data from the models because the metrics are hard to compute from first principles for HMMs with temporally varying transition probabilities.

@@

<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#1.  $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
load("cat1allmod.RData")
df2 <- transform(allmodsumdf,
            distrib=ifelse(grepl("LN",model),"log-Normal","Weibull"),
            turn_angle=ifelse(grepl("VM",model),"yes","no"),
            homog=ifelse(grepl("HMM",model),"homogeneous HMM","time-heterogeneous HMM"))
ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))+
    geom_line()+
    facet_wrap(~ homog)+
    scale_colour_brewer(name="step length\ndistribution",palette="Set1")+
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    scale_linetype(name="von Mises\nturning angle?")+
    labs(y=expression(Delta*"BIC"),
         x="Number of States")+
    geom_hline(yintercept=0,colour="gray")+
    zmargin
@

Figure 1 shows the BIC-optimal number of states for time homogenous models is consistent with \citeauthor{kerk2015hidden}'s results. Furthermore, BIC-optimal number of states does not depend on the model for both time-homogenous and time-inhomogenous within their respective framework, but time-inhomogenous models have the BIC-optimal number of states overall.

<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models">>=
load("cat1sumdat.RData")
ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
    scale_size_continuous( name="Num of Parameters") +
    geom_point(aes(size=parameters))+
    scale_colour_brewer(palette="Dark2")+
    labs(y=expression(paste(Delta, "BIC")),
         x="Number of States") +
    geom_line() + zmargin
@

As a complement, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions (24-hour multiple block transition model) to compare the temporal effects in goodness of fit. Looking at Figure 2, models with temporal heterogenity is better than models without in both FMM and HMM frameworks, but time homogenous HMM models is better than FMM with sinusoidal temporal heterogeneity. Temporal heterogenity is good with good models such as the quadratic and sinuosidal. Multiple block models is good for cats with distinguishable diurnal patterns and hourly model is too complicated.

<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day">>=
  simdf <- simdat %>% dplyr:::select(-obs) %>% group_by(time) %>% summarise_each(funs(mean))
  obsdf <- simdat %>% dplyr:::select(c(obs,time)) %>% filter(!is.na(obs)) %>% group_by(time) %>% summarise_each(funs(mean)) %>% dplyr:::select(obs)
  avgdf <- cbind(simdf,obsdf)
  avgdf2 <- melt(avgdf,'time')
ggplot(avgdf2,aes(x=time,y=value,colour=variable)) + 
    geom_point() +geom_line()+ xlab("Time of Day") + ylab("Step lengths (log10)")+
    scale_x_continuous(breaks=0:23)
@

Figure 3 shows a clear diurnal pattern from the average hourly step lengths from the observed data. Models that assume temporal homogeneity (FMM and HMM) fail to capture the diurnal activity cycle. Models incorporating temporal heterogeneity can capture the observed patterns without the Viterbi algorithm.

<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF">>=
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}

  temp <- simdat %>% dplyr:::select(-time)%>%ldply(.,acffun)
  ggplot(temp,aes(lag,ACF,colour=.id))+
    geom_point()+geom_line()
@

The observed data has strong 24 hour lag autocorrelation (correlation with itself) which suggest temporal heterogeneity. HMM without temporal heterogeneity can capture short-term autocorrelation (without Viterbi algorithm) but is unable to capture lags beyond 7. FMM with temporal heterogeneity can capture lags beyond 12, but not short-term autocorrelations. HMM with temporal heterogeneity can fully capture the autocorrelation pattern of the observed data whereas the FMM completely missed the autocorrelations.

<<state_plots, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Step-length parameters">>=
## translate single fit to data frame of params
trans_pars <- function(fit,npar,Model=NULL) {
    x <- tail(getpars(fit),npar)
    d <- data.frame(log_mean=x[names(x)!="sd"],
                    log_sd=x[names(x)=="sd"])
    if (!is.null(Model)) d$Model <- Model
    d <- arrange(d,log_mean)
    return(d)
}
## get pars for best-fit TH(sin) and HMM models for each cat
get_cat_pars <- function(catnum) {
    load(paste0("cat",catnum,"sumdat.RData"))
    ## figure out
    w <- with(sumdat,which(deltaBIC==min(deltaBIC[model=="HMM + THsin"])))
    optstates <- sumdat$nstates[w]
    ## ASSUME that best fit is het-sin, corresponding homo has +1 states
    sinfit <- get(paste0("fitsin",optstates,"s"))
    homofit <- get(paste0("fithomo",optstates+1,"s"))
    d <- rbind(trans_pars(sinfit,optstates*2,Model="HMM+TH(sin)"),
               trans_pars(homofit,(optstates+1)*2,Model="HMM"))
    return(d)
}
all_cat_pars <- ldply(lme4:::namedList(1,2,14,15),get_cat_pars,.id="catID")
ggplot(all_cat_pars,aes(log_mean,log_sd,color=Model))+
    geom_point()+geom_line() + facet_wrap(~catID,labeller=label_both)
@

@@

For some kinds of model predictions, the time homogeneous transition model can perform as good as the time inhomogeneous transition models in terms of predicting the states by the Viterbi algorithm (the tradition state prediction method) and can easily overlook more complicated models. The Viterbi algorithm solves/predicts for the most likely/probable sequence of movement states based on the observations \citep{zucchini_hidden_2009,langrock_flexible_2012}. This approach allows the time homogeneous transition model's prediction to capture the temporal heterogeneity from the observations. It is useful to predict missing data in the observation sequence but fail to future movements beyond the observations sequence. In general, the time homogeneous transition model gives the broader scope of movement pattern where as the time inhomogeneous transitions models gives a more realistic movement structure and future predictions.

Although the time inhomogeneous transition offer an
objective approach to (1) reduce the number of states to a biologically interpretable level, (2) captures the observational autocorrelation and temporal pattern, (3) reduce the overall complexity as illustrated by our study, several challenges
must be overcome in order to maximize its potential. The current model is based on the reduced univariate step length response emission without fitting turning angles. Although turning angles are strongly correlated with step length, it does not effect the overall prediction on number of states. But movement trajectories predictions, and spatial models still require both step length and turning angles. Furthermore, modelling temporal heterogeneity only reduce the number of predicted states by 1 resulting in 4/5 states might suggest (1) other factors such as spatial components that might also contribute to over predicting number of states, or (2) there really are 1 or 2 more different movement characteristics that are biologically unknown. Despite the challenges, our study provides an important extension in HMM modeling that is not in the seen before and a real life application that it exist and applicable in animal movement ecology.

@@

temp <- vitcat(cat,fithomo3s) %>% dplyr:::select(-time)%>%ldply(.,acffun)
ggplot(temp,aes(lag,ACF,colour=.id))+
geom_point()+geom_line() +theme_bw() +
ggtitle("ACF plot of BIC Selected Models")

@@

<<adj_BIC_comparisons cat2, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models">>=
load("cat2sumdat.RData")
ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
    scale_size_continuous( name="Num of Parameters") +
    geom_point(aes(size=parameters))+
    scale_colour_brewer(palette="Dark2")+
    labs(y=expression(paste(Delta, "BIC")),
         x="Number of States") +
    geom_line() + zmargin
@

>>>>

Clustering time-series data into discrete groups can improve
prediction as well as providing insight into the nature of underlying,
unobservable states of the system. However, temporal variation in the
probability that individuals occupy or move between groups can obscure
such signals. We use finite mixture and hidden Markov models, two
standard clustering techniques, to model high-resolution hourly
movement data from Florida panthers (\emph{Puma concolor coryi}).
Allowing for temporal
heterogeneity in transition probabilities, a straightforward but
rarely explored model extension, resolves some shortcomings of current
modeling frameworks and clarifies the behavioural patterns of
panthers. More generally, incorporating previously neglected
structure in statistical models can allow researchers to more
accurately identify models of appropriate complexity.

@@

\cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as 6 for individuals with a large amount
of available data); for practical reasons, they restricted their
detailed analysis to models with only 3 underlying states.  In
contrast, most studies using HMM have chosen the
number of underlying states \emph{a priori}, typically using either two
\citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,fryxell_multiple_2008}, or three states
\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,kerk2015hidden}. As \cite{kerk2015hidden} comment, and as
we discuss further below, behavioural
repertoires with more than three distinct states are difficult to
interpret --- one possible reason that other authors have not adopted
\citeauthor{kerk2015hidden}'s model-based approach to
identifying the number of latent states.

Our second goal, therefore, is to explore whether
\citeauthor{kerk2015hidden}'s results might be driven at
least in part by structural problems with their statistical
model, i.e. the
assumption of temporally homogeneous behaviour.  When large data sets
are available, information-theoretic model selection methods will
typically choose complex, highly parameterized models; when there is
only one way in which models can become more complex (e.g. by
increasing the number of latent states), complexity that is
present in the data but not accounted for in the model (e.g. spatial
or temporal heterogeneity) can be misidentified as other forms of
complexity.  We predict that increasing volumes of data will
increasingly lead researchers who are accustomed to fitting 
small models to sparse data into such traps.  We examine whether
allowing for diurnal variation in the Florida panther data leads to
selection of models with smaller numbers of latent states; we also
fit models to simulated data with varying numbers of latent states
and degrees of temporal heterogeneity to test our conjecture that
heterogeneity can be misidentified as behavioural complexity.

@@

\cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
an extension of HMM that permits explicit modelling of dwell times 
\citep{langrock_flexible_2012}, considering both Poisson and
negative binomial distributions for dwell times.  As shown 
by \cite{kerk2015hidden} (Figure S3b, top row, middle panel),
the estimated shape parameter of the negative binomial dwell time
distribution was typically close to 1 ($\approx 0.4-1.6$; confidence
intervals were not calculated), implying that a geometric distribution
(i.e., negative binomial with shape=1) might be adequate. In turn,
this suggests that we might not lose much accuracy by reverting to 
a simpler HMM framework, which corresponds
to making precisely this assumption.

@@

\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent states.
In general, BIC values decreased as the number of states
increased from 
3 to 6 states,
suggesting that the 6-state model was
favoured statistically; however,
the authors used 3-state models in most
of their analyses for ease of biological interpretation.
We follow \cite{kerk2015hidden} in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the ``true'' number 
of underlying movement states.  
\todo{it occurs to me that we haven't really included any
information about simulation testing in this version!
We should probably add a short section; if don't, then we
should delete following sentence}
Using BIC also simplifies 
evaluation of model selection procedures; 
it is easier to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\citep{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \citep{BurnhamAnderson1998};
we suspect that the qualitative conclusions stated here
for BIC-optimality will carry over to analyses using AIC instead.

@@

In a HMM, the joint likelihood of \emph{emissions} (i.e., direct observations) $\mathbf Y = \obs_{1},..., \obs_{T}$ and a hidden state sequence $\mathbf{Z}, \state_{t} \in \{1,...,n\}, t=1,...,T$, given model parameters {$\boldsymbol \theta$} and covariates $\mathbf{X}_{1:T} =\cov_{1},..., \cov_{T}$, can be written as:
\begin{equation}
\begin{split}
P(\mathbf{Y}_{1:T},\mathbf{Z}_{1:T}|\boldsymbol \theta,\mathbf{X}_{1:T}) & =
P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \cdot \\
&  \quad \prod\limits_{k=2}^{T}P(\state_{k} | \state_{k-1}, \cov_{k})P(\obs_{k} | \state_{k},\cov_{k})
\end{split}
\label{eq:HMMlik}
\end{equation}

The emissions $\obs_i$ are boldfaced to denote that we may have a vector of observations at each time point (e.g., step length and turning angle).
The model contains three distinct components:

\begin{description}
\item[Initial probability] $P(\state_{1} = i | \cov_{1}) P(\obs_{1} | \state_{1}, \cov_{1} )$: the probability of state $i$ at time $t=1$ where the covariate is $\cov_{1}$, times the vector of observations $\obs_{1}$ conditioned on covariates $\cov_{1}$ and state $\state_{1}$.

@@

\item[Emission probability] $P(\obs_{k} | \state_{k},\cov_{k})$: a vector of observations $\obs_{k}$ conditioned on covariates $\cov_{k}$ at state $\state_{k}$ at time $t=k$.  
\end{description}

\todo{maybe give the formula for number of parameters (constrained) for each model type here?}

@@

We considered four different transition models for diurnal variation in behaviour, incorporating hour-of-day as a covariate following the general approach of \cite{morales_extracting_2004} of incorporating covariate dependence in the transition matrix.

\begin{description}
\item[Hourly model] \todo{describe hourly model. Explain that it's included for comparative purposes, i.e. not really practical.  Explain that we only fitted up to 4 states (because too many parameters)}
\item[Multiple block transition] Here we assume piecewise-constant transition probabilities. The transition probability $\pi_{ij}$ is a function of time (hour of day), where it is assigned to one of $M$ different time blocks:
$$
\eta_{ij}(t) = \sum_{m=1}^{M}a_{ijm} \delta_{m=t}
$$
where $a_{ijm}$ are parameters, and $\delta_{m=t}$ is a Kronecker delta ($\delta_{m=t}=1$ for the time block at the corresponding time $t$, and 0 otherwise).  
\item[Quadratic transition model] We assume the elements of the linear predictor
are quadratic functions of hour. The quadratic model is not diurnally continuous, i.e. there is no constraint that forces $\eta_{ij}(0)=\eta_{ij}(24)$; imposing a diurnal continuity constraint would collapse the model to a constant.
$$
\eta_{ij}(t) = b_{ij1}+b_{ij2}\left(\frac{t}{24}\right)+b_{ij3}\left(\frac{t}{24}\right)^{2}
$$
\item[Sinusoidal transition model] A sinusoidal model with a period of 24 hours is identical in complexity to the quadratic model, but automatically satisfies the diurnal continuity constraint.
$$
\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos\left(\frac{2\pi t}{24}\right)+b_{ij3}\sin\left(\frac{2\pi t}{24}\right)
$$
\end{description}

Other periodic functions, such as Fourier series (the sinusoidal transition model augmented by additional sinusoidal components at higher frequencies) or periodic splines\todo{citation?}, could be useful directions for future exploration.

Model complexity and the number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity (Supplementary material, Table 1).

@@

We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers with the most data ($>  $ observations). (1) Comparing BICs to the optimal-BIC model within each type of transition complexity ($\dbic$ = BIC - min(BIC)) assesses the overall goodness of
fit of each model type.
(2) Comparing average step-length by hour of day for the observed data and for data simulated from the models shows how well a particular class of models can capture the diurnal variation in behaviour.
(3) Comparing temporal autocorrelations for the observed data and for data simulated from the models shows how well a particular class of models
captures serial correlation at both short and long scales. 
We used simulations to predict hourly step length and ACF 
because, while the computation is reasonably straightforward for FMMs, and manageable for homogeneous HMMs, the interaction between the geometric dwell time within each state and the temporally varying interaction probabilities makes it unreasonably complex.
\todo{need to explain here, if we haven't elsewhere, that using the Viterbi algorithm to simulate double-counts the data ...}

@@

<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#1.  $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
L <- load("cat1allmod.RData")
df2 <- transform(allmodsumdf,
            distrib=ifelse(grepl("LN",model),"log-Normal","Weibull"),
            turn_angle=ifelse(grepl("VM",model),"yes","no"),
            homog=ifelse(grepl("HMM",model),"homogeneous HMM","time-heterogeneous HMM"))
ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))+
    geom_line()+
    facet_wrap(~ homog)+
    scale_colour_brewer(name="step length\ndistribution",palette="Set1")+
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    scale_linetype(name="von Mises\nturning angle?")+
    labs(y=expression(Delta*"BIC"),
         x="Number of States")+
    geom_hline(yintercept=0,colour="gray")+
    zmargin
@

Figure~\ref{fig:BICred_plot} shows the BIC-optimal number of states for time homogenous models is consistent with \citeauthor{kerk2015hidden}'s results. Furthermore, BIC-optimal number of states does not depend on the model for both time-homogenous and time-inhomogenous within their respective framework, but time-inhomogenous models have the BIC-optimal number of states overall.\todo{do we need to run some models with 7 states to show that $\dbic$ eventually increases?}

<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall adjusted BIC comparisons among models",fig.width=8>>=
L <- load("cat1sumdat.RData")
ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
    scale_size_continuous( name="# of parameters") +
    scale_x_continuous(expand=c(0,0.5),breaks=2:6)+
    geom_point(aes(size=parameters))+
    scale_colour_brewer(palette="Dark2")+
    labs(y=expression(paste(Delta, "BIC")),
         x="Number of States") +
    geom_line() + zmargin
@

As a complement, we also fitted FMM; FMM with sinusoidal variation in state occupancy probabilities; and HMM with hourly transitions (24-hour multiple block transition model) to compare the temporal effects in goodness of fit. Figure~\ref{fig:adj_BIC_comparisons} shows that models with temporal heterogenity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogenous HMM models are better than FMM with sinusoidal temporal heterogeneity. Turning to the temporally heterogeneous HMMs (right panel), we see that the model with different transition probabilities for each hour of the day (HMM + THhourly) is overparameterized; it underperforms homogeneous HMM with even 3 states, and gets much worse with 4 states. The multiple-block model approximately matches the homogeneous HMM, although it gives the BIC-optimal number of states as 4, in contrast to 6 for the homogeneous HMM.  Finally, the quadratic and sinusoidal models are considerably better than any other models tested; they both give the BIC-optimal number of states as 5, and they have similar goodness of fit.
However, this similarity is somewhat overstated due to the very large variation in BIC (over thousands of units) across the full range of models; there is a difference of approximately 80 BIC units, which would
normally be interpreted as an enormous difference in goodness of fit,
between the sinusoidal and quadratic models (both of which have 90 
parameters).

<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day. \\textbf{todo: leave out hmm quad for clarity? fix factor labels; make y axis log-scale rather than showing log-step-length; make colours consistent with prev fig (may need scalecolourmanual [need underscores] with RColorBrewer::brewer.pal(10,'Dark2') or some such. Maybe set obs as a wide, pointless, semitransparent line?})",fig.width=8>>=
simdf <- simdat %>% dplyr::select(-obs) %>%
    group_by(time) %>% summarise_each(funs(mean))
obsdf <- simdat %>% dplyr::select(c(obs,time)) %>%
    filter(!is.na(obs)) %>% group_by(time) %>%
    summarise_each(funs(mean))

avgdf2 <- melt(simdf,'time')
ggplot(avgdf2,aes(x=time,y=value,colour=variable)) + 
    geom_point() +geom_line()+ xlab("Time of Day") +
    ylab("Step lengths (log10)")+
    scale_x_continuous(breaks=0:23)+
    geom_line(data=obsdf,
              aes(y=obs),lwd=2.5,alpha=0.3,colour="black")
@

Figure~\ref{fig:avg_step_length_by_time} shows a clear diurnal pattern from the average hourly step lengths from the observed data. As expected, temporally homogeneous models (FMM and HMM) predict the same mean step length regardless of time of day, failing to capture the diurnal activity cycle. All of the models incorporating temporal heterogeneity, including the temporally heterogeneous FMM, can capture the observed patterns. However, the block model does markedly worse than the other temporal models, and the (overparameterized) hourly model does better than any other model at capturing the early-evening peak (but worse at capturing the mid-day trough).

<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF \\textbf{fix similarly to Fig 3}">>=
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}
temp <- simdat %>% dplyr::select(-time)%>%ldply(.,acffun)
  ggplot(temp,aes(lag,ACF,colour=.id))+
    geom_point()+geom_line()
@

The observed data has strong 24 hour lag autocorrelation (correlation with itself) which suggest temporal heterogeneity (Figure~\ref{fig:acf_plot}). HMM without temporal heterogeneity can capture short-term autocorrelation (without Viterbi algorithm) but is unable to capture lags beyond 7. FMM with temporal heterogeneity can capture lags beyond 12, but not short-term autocorrelations. HMM with temporal heterogeneity can fully capture the autocorrelation pattern of the observed data whereas the FMM completely missed the autocorrelations.

<<get_cat_pars, warning=FALSE, message=FALSE, echo=FALSE, cache=TRUE>>=
## translate single fit to data frame of params
trans_pars <- function(fit,npar,Model=NULL) {
    x <- tail(getpars(fit),npar)
    d <- data.frame(log_mean=x[names(x)!="sd"],
                    log_sd=x[names(x)=="sd"])
    if (!is.null(Model)) d$Model <- Model
    d <- arrange(d,log_mean)
    return(d)
}
## get pars for best-fit TH(sin) and HMM models for each cat
get_cat_pars <- function(catnum) {
    L <- load(paste0("cat",catnum,"sumdat.RData"))
    ## figure out
    w <- with(sumdat,which(deltaBIC==min(deltaBIC[model=="HMM + THsin"])))
    optstates <- sumdat$nstates[w]
    ## ASSUME that best fit is het-sin, corresponding homo has +1 states
    sinfit <- get(paste0("fitsin",optstates,"s"))
    homofit <- get(paste0("fithomo",optstates+1,"s"))
    d <- rbind(trans_pars(sinfit,optstates*2,Model="HMM+TH(sin)"),
               trans_pars(homofit,(optstates+1)*2,Model="HMM"))
    return(d)
}
all_cat_pars <- ldply(lme4:::namedList(1,2,14,15),get_cat_pars,.id="catID")
@

<<state_plots, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Step-length parameters">>=
ggplot(all_cat_pars,aes(log_mean,log_sd,color=Model))+
    geom_point()+geom_line() + facet_wrap(~catID,labeller=label_both)
@

TODO: add ellipses?

@@

For some kinds of model predictions, the time homogeneous transition model can perform as good as the time inhomogeneous transition models in terms of predicting the states by the Viterbi algorithm (the tradition state prediction method) and can easily overlook more complicated models. The Viterbi algorithm solves/predicts for the most likely/probable sequence of movement states based on the observations \citep{zucchini_hidden_2009,langrock_flexible_2012}. This approach allows the time homogeneous transition model's prediction to capture the temporal heterogeneity from the observations. It is useful to predict missing data in the observation sequence, but because it is conditional on the observed values, it cannot predict future movements beyond the observation sequence. In general, the time homogeneous transition model gives the broader scope of movement pattern where as the time inhomogeneous transitions models gives a more realistic movement structure and future predictions.

Although the time inhomogeneous transitions offer an
objective approach to (1) reduce the number of states to a biologically interpretable level, (2) captures the observational autocorrelation and temporal pattern, (3) reduce the overall complexity as illustrated by our study, several challenges
must be overcome in order to maximize its potential. The current model is based on the reduced univariate step length response emission without fitting turning angles. Although turning angles are strongly correlated with step length, it does not effect the overall prediction on number of states. But movement trajectories predictions, and spatial models still require both step length and turning angles. Furthermore, modelling temporal heterogeneity only reduce the number of predicted states by 1 resulting in 4/5 states might suggest (1) other factors such as spatial components that might also contribute to over predicting number of states, or (2) there really are 1 or 2 more different movement characteristics that are biologically unknown. Despite the challenges, our study provides an important extension in HMM modeling that is not in the seen before and a real life application that it exist and applicable in animal movement ecology.

@@

temp <- vitcat(cat,fithomo3s) %>% dplyr::select(-time)%>%ldply(.,acffun)
ggplot(temp,aes(lag,ACF,colour=.id))+
geom_point()+geom_line() +theme_bw() +
ggtitle("ACF plot of BIC Selected Models")

@@

<<adj_BIC_comparisons_cat2, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall adjusted BIC comparisons between different HMM models">>=
load("cat2sumdat.RData")
ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
    scale_size_continuous( name="Num of Parameters") +
    geom_point(aes(size=parameters))+
    scale_colour_brewer(palette="Dark2")+
    labs(y=expression(paste(Delta, "BIC")),
         x="Number of States") +
    geom_line() + zmargin
@

<<adj_BIC_comparisons_cat14, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models">>=
load("cat14sumdat.RData")
ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
    scale_size_continuous( name="Num of Parameters") +
    geom_point(aes(size=parameters))+
    scale_colour_brewer(palette="Dark2")+
    labs(y=expression(paste(Delta, "BIC")),
         x="Number of States") +
    geom_line() + zmargin
@
